---
title: "quallmer audit trail"
date: "2026-06-17"
format:
  html:
    toc: true
    toc-depth: 3
---

This audit trail documents the complete workflow following Lincoln and Guba's (1985)
framework for establishing trustworthiness in qualitative research.

## Trail summary

- **Number of runs:** 4
- **Complete:** Yes

### System information

- **quallmer version:** 0.4.0
- **ellmer version:** 0.4.0
- **R version:** 4.5.3

## Instrument development

Codebooks used in this analysis:

### Ideological Scaling

**Instructions:**

> Read each text carefully. Place the text on a -5 to +5 scale
    for the inclusive-exclusive ideological dimension.

    INCLUSIVE language (-5): Emphasizes equal rights, diversity, pluralism,
    and protection of minorities.

    EXCLUSIVE language (+5): Emphasizes exclusion of groups, national homogeneity,
    and restricting rights.

    Score 0 = neutral or mixed rhetoric.

**Schema:**

```
<ellmer::TypeObject>
 @ description          : NULL
 @ required             : logi TRUE
 @ properties           :List of 2
 .. $ score      : <ellmer::TypeBasic>
 ..  ..@ description: chr "Ideological position (-5 = inclusive, +5 = exclusive)"
 ..  ..@ required   : logi TRUE
 ..  ..@ type       : chr "integer"
 .. $ explanation: <ellmer::TypeBasic>
 ..  ..@ description: chr "Brief justification for the assigned score, referring to specific text elements"
 ..  ..@ required   : logi TRUE
 ..  ..@ type       : chr "string"
 @ additional_properties: logi FALSE
```

## Process notes

Chronological record of all coding runs:

### 1. run1_ideology

**Parent:** None (original run)
**Timestamp:** 2026-06-17 20:12:33
**Model:** openai/gpt-4o-mini
**Codebook:** Ideological Scaling
**Units coded:** 5

**Call:**

```{r eval=FALSE}
qlm_code(x = inaugural_texts, codebook = ideology_codebook, model = "openai/gpt-4o-mini", 
    name = "run1_ideology")
```

### 2. run2_same_settings

**Parent:** run1_ideology
**Timestamp:** 2026-06-17 20:29:50
**Model:** openai/gpt-4o-mini
**Codebook:** Ideological Scaling
**Units coded:** 5

**Call:**

```{r eval=FALSE}
qlm_replicate(x = coded_run1, name = "run2_same_settings")
```

### 3. run3_high_temp

**Parent:** run1_ideology
**Timestamp:** 2026-06-17 20:30:26
**Model:** openai/gpt-4o-mini
**Codebook:** Ideological Scaling
**Units coded:** 5

**Call:**

```{r eval=FALSE}
qlm_replicate(x = coded_run1, params = params(temperature = 0.9), 
    name = "run3_high_temp")
```

### 4. run4_llama

**Parent:** run1_ideology
**Timestamp:** 2026-06-17 20:31:52
**Model:** ollama/llama3.2:1b
**Codebook:** Ideological Scaling
**Units coded:** 5

**Call:**

```{r eval=FALSE}
qlm_replicate(x = coded_run1, model = "ollama/llama3.2:1b", name = "run4_llama")
```

## Raw data and coded results

The complete coded data for each run is stored in the RDS file.
Load with `readRDS()` to access:

```r
trail <- readRDS("path/to/trail.rds")
trail$runs$run_name$data
# Coded results for a specific run
```

**run1_ideology:** 5 units, 2 variables
**run2_same_settings:** 5 units, 2 variables
**run3_high_temp:** 5 units, 2 variables
**run4_llama:** 5 units, 2 variables

## Replication

This section provides instructions and code to replicate the analysis.

### Environment setup

Install required packages:

```r
# Install quallmer (if not already installed)
# install.packages("pak")
# pak::pak("quallmer/quallmer")

library(quallmer)
```

Configure API credentials for the models used:

```r
# OpenAI: Set OPENAI_API_KEY environment variable
# Sys.setenv(OPENAI_API_KEY = "your-api-key")
# Ollama: Ensure Ollama is running locally
# See: https://ollama.ai/
```

### Loading the trail

Load the saved trail to access codebooks, data, and metadata:

```r
trail <- readRDS("ideology_analysis.rds")
```

### Replication code

The following code replicates each coding run using the stored parameters.

#### Replicate: run1_ideology

```r
# Get codebook from trail
codebook_run1_ideology <- trail$runs$run1_ideology$codebook

# Load your input data (same structure as original)
# texts <- your_data$text_column

coded_run1_ideology <- qlm_code(
  texts,
  codebook_run1_ideology,
  model = "openai/gpt-4o-mini",
  name = "run1_ideology"
)
```

#### Replicate: run2_same_settings

```r
# Get codebook from trail
codebook_run2_same_settings <- trail$runs$run2_same_settings$codebook

# Load your input data (same structure as original)
# texts <- your_data$text_column

coded_run2_same_settings <- qlm_code(
  texts,
  codebook_run2_same_settings,
  model = "openai/gpt-4o-mini",
  name = "run2_same_settings"
)
```

#### Replicate: run3_high_temp

```r
# Get codebook from trail
codebook_run3_high_temp <- trail$runs$run3_high_temp$codebook

# Load your input data (same structure as original)
# texts <- your_data$text_column

coded_run3_high_temp <- qlm_code(
  texts,
  codebook_run3_high_temp,
  model = "openai/gpt-4o-mini",
  name = "run3_high_temp"
)
```

#### Replicate: run4_llama

```r
# Get codebook from trail
codebook_run4_llama <- trail$runs$run4_llama$codebook

# Load your input data (same structure as original)
# texts <- your_data$text_column

coded_run4_llama <- qlm_code(
  texts,
  codebook_run4_llama,
  model = "ollama/llama3.2:1b",
  name = "run4_llama"
)
```

### Note on reproducibility

LLM outputs are inherently stochastic. To improve reproducibility:

- Use `temperature = 0` for more deterministic outputs
- Set a random seed where supported by the provider
- Document the exact model version (models are updated over time)
- Compare results across multiple runs using `qlm_replicate()`

## Reference

Lincoln, Y. S., & Guba, E. G. (1985). *Naturalistic Inquiry*. Sage.

