---
title: "The quallmer audit trail"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The quallmer audit trail}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

The `qlm_trail()` function creates audit trails following Lincoln and Guba's (1985) framework for establishing trustworthiness in qualitative research. It captures the complete decision history of your coding workflow.

## The audit trail concept

Lincoln and Guba (1985, p. 319-320) describe six categories of audit trail materials for qualitative research:

1. **Raw data**: Original texts
2. **Data reduction products**: Summaries, coded results
3. **Data reconstruction products**: Themes, comparisons, findings
4. **Process notes**: Methodological decisions, procedures
5. **Materials relating to intentions**: Research proposals, notes
6. **Instrument development**: Codebooks, protocols

quallmer automatically captures these materials as you work:

| Component | What quallmer stores |
|-----------|---------------------|
| Raw data | Original texts in coded objects |
| Data reduction | Coded results from each run |
| Data reconstruction | Comparison and validation results |
| Process notes | Model, temperature, timestamps, user notes |
| Intentions | The R function calls you made |
| Instruments | Codebook instructions and schema |

## How it works

Every quallmer object carries metadata about how it was created. The trail system connects these objects to show your complete workflow:

```
qlm_code()       →  Creates coded object with run metadata
      ↓
qlm_backfill()   →  Completes the run; each pass is recorded on the same object
      ↓
qlm_replicate()  →  Creates new object linked to parent
      ↓
qlm_compare()    →  Creates comparison linked to inputs
      ↓
qlm_trail()      →  Extracts and organizes all metadata
```

## Testing branches

After initial coding with `qlm_code()`, the workflow branches into different types of quality assessment:

```
                           ┌─── a) Robustness (test-retest reliability)
                           │    qlm_replicate() → qlm_compare()
                           │
qlm_code() ────────────────┼─── b) Reliability (test-test across models)
                           │    qlm_replicate(<alter settings>) → qlm_compare()
                           │
                           └─── c) Accuracy (vs. gold standard)
                                qlm_validate()
```

**a) Robustness via repetition**: Run the same model and settings multiple times to establish test-retest reliability. Use `qlm_replicate()` with identical settings, then `qlm_compare()` to measure consistency.

**b) Reliability via different models**: Compare results across different LLMs or temperatures to establish test-test reliability. Use `qlm_replicate()` with altered settings (e.g., different model, temperature), then `qlm_compare()` to assess agreement.

**c) Accuracy versus a gold standard**: If human-coded gold standard data exists, use `qlm_validate()` to compute accuracy metrics (precision, recall, F1, etc.).

`qlm_trail()` collects all these branches into a single trail object, providing a complete record of what was tested and how the results compare.

## Creating a trail

Pass any quallmer objects to `qlm_trail()`:

```{r}
library(quallmer)

# After your analysis...
coded1 <- qlm_code(
  texts, codebook,
  model = "openai/gpt-4o",
  name = "gpt4o",
  notes = "Initial coding with GPT-4o"
)

coded2 <- qlm_replicate(
  coded1,
  model = "anthropic/claude-sonnet-4",
  name = "claude",
  notes = "Replication to check model agreement"
)

comparison <- qlm_compare(coded1, coded2, by = sentiment)

# Create the audit trail
trail <- qlm_trail(coded1, coded2, comparison)
trail
```

The print output shows:

- Run names and parent relationships
- Timestamps and model information
- Backfill passes, and any other model they used, so a run completed by two instruments is disclosed as one
- Notes documenting why each run was performed
- Comparison/validation summaries
- Whether the chain is complete

## Saving the trail

Add a `path` argument to save permanent files:

```{r}
qlm_trail(coded1, coded2, comparison, path = "my_analysis")
```
This creates:

- **my_analysis.rds**: Complete R object with all coded data
- **my_analysis.qmd**: Quarto document for human-readable documentation

## The Quarto report

The generated `.qmd` file contains:

1. **Trail summary**: Number of runs, completeness status
2. **Instrument development**: All codebooks used with instructions
3. **Process notes**: Chronological record of each run
4. **Data reconstruction**: Comparison and validation results
5. **Replication section**: Code to reproduce the analysis

Render it with:

```{r}
quarto::quarto_render("my_analysis.qmd")
```

## Loading saved trails

Reload a trail to access the data:

```{r}
trail <- readRDS("my_analysis.rds")

# Access runs
names(trail$runs)

# Get coded data from a run
trail$runs$gpt4o$data

# Check run metadata
trail$runs$gpt4o$metadata$timestamp
trail$runs$gpt4o$chat_args$name
```

## Best practices

1. **Name your runs**: Use the `name` parameter for readable trails
2. **Add notes**: Use the `notes` parameter to document why each run was performed
3. **Include all objects**: Pass every relevant object to `qlm_trail()` for a complete chain
4. **Save early**: Create trail files before sharing or publishing results

## Reference

Lincoln, Y. S., & Guba, E. G. (1985). *Naturalistic Inquiry*. Sage.
