---
title: "The quallmer workflow"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

This tutorial provides a quick overview of the quallmer workflow for LLM-assisted qualitative coding. We'll use US presidential inaugural addresses to demonstrate the core functions.

## The workflow at a glance

1. **Define** your codebook with `qlm_codebook()`
2. **Code** your data with `qlm_code()`, and complete a run that came back with failures using `qlm_backfill()`
3. **Replicate** with different settings using `qlm_replicate()`
4. **Compare** results with `qlm_compare()`
5. **Document** everything with `qlm_trail()`

## Sample data

We'll use the inaugural addresses from the quanteda package:

```{r}
library(quallmer)
library(quanteda)

# Get last 5 inaugural addresses
texts <- tail(data_corpus_inaugural, 5)
texts
```

## Step 1: Define a codebook

A codebook specifies what the LLM should extract from your texts:
```{r}
my_codebook <- qlm_codebook(
  name = "Tone analysis",
  instructions = "Classify the overall tone of this political speech.",
  schema = type_object(
    tone = type_enum(
      values = c("optimistic", "cautious", "urgent"),
      description = "The dominant emotional tone"
    ),
    confidence = type_integer("Confidence in classification from 1-5")
  )
)

my_codebook
```

> **Tip:** Need to split texts into smaller units first? Use `qlm_segment()` to segment texts into thematic or conceptual units (e.g., quasi-sentences, aspects, or topics) before coding. For an example, see here: [Text segmentation](https://quallmer.github.io/quallmer/articles/pkgdown/examples/example_segmentation.html).

## Step 2: Code your data

Apply the codebook to your texts using an LLM:

```{r}
coded <- qlm_code(
  texts,
  my_codebook,
  model = "openai/gpt-4o-mini",
  name = "gpt4o_mini"
)

coded
```

### When a run comes back incomplete

A run over a real corpus rarely comes back complete. Requests time out or hit a rate limit, a provider refuses a text on one pass and codes it on the next, an endpoint accepts the schema and then returns something that does not fit it. The units that failed sit in the object as rows of `NA`, and `qlm_failures()` lists them with the reason:

```{r}
qlm_failures(coded)
```

quallmer deals with failures at three points, and it helps to keep them apart.

**Retries happen during the run, one request at a time, in two layers.** The inner layer is ellmer's: every request quallmer sends is retried at the transport level if a rate limit or a server error turns it away, a few times with a pause between, on every provider and every path; `options(ellmer_max_tries = )` sets the count. The outer layer is quallmer's, and exists on the JSON path only: when a request comes back unusable after ellmer's tries, whether empty, unparsable, refused, not conforming to the codebook's schema, or still failing on transport, `qlm_code()` sends the unit again, up to `json_retries` more times, and each of those requests gets ellmer's tries afresh. On the structured path there is no outer layer: a request is tried by ellmer and then stands, and the only second chance is that `structured = "auto"` falls back to the JSON path when the structured call fails as a whole. A unit still failed after all of this is what `qlm_failures()` shows.

**Backfilling happens after the run, on the object it produced.** `qlm_backfill()` looks up the units that failed, re-codes just those with the run's own model and settings, and merges what comes back; everything that succeeded the first time is left exactly as it was. Because it works on the finished object, you can run it later, look at the failures first, or give it a different `model` for units the original consistently refuses or cannot fit in its context window. Each pass is recorded in the object's metadata, and `print()` and the audit trail say what was backfilled and by which model, so a result coded by two instruments is disclosed as one.

```{r}
filled <- qlm_backfill(coded)
qlm_failures(filled)
```

The two are for different failures. Retries catch what a moment's wait fixes, in the middle of the run. Backfilling recovers what is left when the run is over, which for a corpus of any size is usually something. Two kinds of failure a backfill leaves alone, because re-sending the same request cannot change the outcome: a text the provider rejected as longer than the model's context window, and a response cut off at `max_tokens`. Those are retried only with a different `model`, or for a truncated response a higher `params(max_tokens = )`. Content refusals are retried, since the same text is refused on one pass and coded on the next.

| Failure | Stage and mechanism | Setting or action |
|---|---|---|
| Rate limit, timeout, or transient server error | ellmer retries the transport request, on every path and for every request quallmer sends | `options(ellmer_max_tries = n)` |
| Unusable response on the JSON path after ellmer's tries: empty, malformed, refused, schema-invalid, or still failing on transport | quallmer sends the unit again during the run, each request with ellmer's tries afresh | `json_retries = n` |
| Anything still failed after the run | quallmer makes another pass over the failed units | `backfill = n` in `qlm_code()`, or `qlm_backfill(..., passes = n)` |
| Input exceeds the model's context window | Backfill with a larger-context model | `qlm_backfill(..., model = )` |
| Output was cut off at `max_tokens` | Backfill with a higher output limit | `qlm_backfill(..., params = ellmer::params(max_tokens = ))` |

Two settings shape how much there is to backfill. By default a run stops sending new requests after the first failure and returns what it has, which is ellmer's `on_error = "return"`; pass `on_error = "continue"` to have it attempt every unit and leave the failures for a backfill. And `qlm_code(backfill = 2)` runs the passes in the same call, if you would rather not make them a separate step. Either way, a replication made with `qlm_replicate()` replays the passes recorded on its parent, so both are complete on the same terms.

## Step 3: Replicate with a different model

Test reliability by coding again with a different model:

```{r}
coded2 <- qlm_replicate(
  coded,
  model = "openai/gpt-4o",
  name = "gpt4o"
)

coded2
```

## Step 4: Compare results

Assess inter-rater reliability between the two coding runs:

```{r}
comparison <- qlm_compare(coded, coded2)
comparison
```

If you have gold standard human coding, you can also validate against it with `qlm_validate()`.

## Step 5: Create an audit trail

Document your complete workflow, including models, parameters, and results and a Quarto report with replication instructions:

```{r}
# View the trail
trail <- qlm_trail(coded, coded2, comparison)
trail

# Save trail and generate report
qlm_trail(coded, coded2, comparison, path = "my_analysis")
# Creates: my_analysis.rds, my_analysis.qmd
```

This workflow provides a structured approach to leveraging LLMs for qualitative coding with transparency as well as full traceability and the ability to replicate and validate your analyses. 
