---
title: "Example: Quasi-sentence segmentation"
subtitle: "Segmenting party manifestos using Manifesto Project rules with `qlm_segment()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example: Quasi-sentence segmentation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5
)
```

The [Manifesto Project](https://manifesto-corpus.wzb.eu/) codes party manifestos by segmenting them into *quasi-sentences* — the smallest units carrying a single, complete political statement. A quasi-sentence is at minimum one natural sentence, but a sentence may be split into multiple quasi-sentences when it contains two or more unique arguments. Applying these rules requires judgment: when are two clauses genuinely independent claims versus elaborations of a single point? This is exactly the kind of nuanced, instruction-following task where LLMs can serve as a scalable alternative to human coding.

This article uses `qlm_segment()` to apply the Manifesto Project quasi-sentence rules to the 1972 New Zealand National Party election manifesto, and evaluates how faithfully the model follows the handbook instructions.

## Packages

```{r packages, message=FALSE, warning=FALSE}
library(quallmer)
library(quanteda)
library(dplyr)
library(ggplot2)
```

## The data

The manifesto is stored as a plain-text file. We read it in and wrap it in a named character vector so the document identifier appears in the output corpus.

```{r load-text}
manifesto_path <- "data/quasi-sentences/NZ_NP_1972.txt"
manifesto_text <- paste(readLines(manifesto_path), collapse = "\n")
manifesto <- c(NZ_NP_1972 = manifesto_text)
cat(substr(manifesto_text, 1, 500), "...\n")
```

## The codebook

The codebook instructions are the verbatim text of section 3.2 ("Unitising — Cutting Text into Quasi-Sentences") from the Manifesto Project coding handbook (Burst et al., 2021), including the rules for when to cut and when not to cut, the worked example, and the expected output. We load the file at runtime and append a short tail telling the model how to format its output.

```{r codebook}
qs_instructions <- paste(
  readLines("data/quasi-sentences/instructions.txt"),
  collapse = "\n"
)

cb_qs <- qlm_codebook(
  name = "Manifesto Project quasi-sentence segmentation",
  instructions = paste(
    qs_instructions,
    "",
    "Return every quasi-sentence in document order.",
    "Each returned 'text' must be verbatim text copied exactly from the input.",
    "Section headers (capitalised lines without end-of-sentence punctuation)",
    "should be joined to the first sentence that follows them.",
    "Mark whether each quasi-sentence is a COMPLETE natural sentence or a FRAGMENT",
    "cut from a larger natural sentence.",
    sep = "\n"
  ),
  schema = ellmer::type_object(
    sentence_type = ellmer::type_enum(
      c("complete", "fragment"),
      description = paste(
        "Whether this quasi-sentence is a complete natural sentence ('complete')",
        "or a fragment cut from a larger natural sentence that was split ('fragment')"
      )
    ),
    reason = ellmer::type_string("Rule governing the segmentation decision.")
  ),
  role = "You are an expert political science coder trained in the Manifesto Project methodology."
)

cb_qs
```

## Segmenting the manifesto

```{r segment-manifesto, eval=FALSE}
segs_manifesto <- qlm_segment(
  manifesto,
  codebook = cb_qs,
  model    = "anthropic/claude-opus-4-6",
  name     = "Claude Opus 4.6"
)
saveRDS(segs_manifesto, "data/segs_manifesto_nz.rds")
```

```{r load-segs-manifesto, echo=FALSE}
segs_manifesto <- readRDS("data/segs_manifesto_nz.rds")
```

## Results

### All quasi-sentences

The model produced `r ndoc(segs_manifesto)` quasi-sentences from the manifesto. The full segmentation is shown below. Each quasi-sentence is numbered, labelled by type (`complete` or `fragment`), and displayed on its own line.

```{r segment-table, results='asis'}
dv <- docvars(segs_manifesto) |>
  mutate(text = as.character(segs_manifesto))

cat(sprintf("**%d.** _%s_\n> %s\n\n", dv$segid, dv$sentence_type, dv$text))
```

### Complete sentences versus fragments

Quasi-sentences labelled `fragment` were cut from a natural sentence that contained more than one unique argument. The split rate provides a rough calibration check: a very low rate suggests the model is treating every sentence as a single unit; a very high rate suggests over-splitting.

```{r split-summary}
dv |>
  count(sentence_type) |>
  mutate(pct = round(100 * n / sum(n), 1)) |>
  knitr::kable(
    col.names = c("Sentence type", "Count", "%"),
    caption   = "Quasi-sentence types"
  )
```

As an additional sanity check, fragments cut from natural sentences should tend to be shorter than complete sentences. The distribution of segment lengths (in characters) confirms this pattern:

```{r plot-lengths, fig.height=4}
dv |>
  mutate(
    nchar         = nchar(text),
    sentence_type = factor(sentence_type, levels = c("complete", "fragment"))
  ) |>
  ggplot(aes(x = nchar, fill = sentence_type)) +
  geom_histogram(binwidth = 30, colour = "white", linewidth = 0.2) +
  facet_wrap(~sentence_type, ncol = 1, scales = "free_y") +
  scale_fill_manual(values = c(complete = "#5cb85c", fragment = "#d9534f"), guide = "none") +
  labs(
    x        = "Quasi-sentence length (characters)",
    y        = "Count",
    title    = "Length distribution by quasi-sentence type",
    subtitle = "Fragments are typically shorter than complete sentences"
  ) +
  theme_minimal()
```

### Coding decisions: why sentences were split

The table below shows every quasi-sentence the model labelled as a `fragment` — a piece cut from a natural sentence that contained more than one unique argument — together with the preceding quasi-sentence for context and the model's cited reason for the split. These are the cases most worth human review: each fragment should represent a genuinely distinct political claim.

```{r split-decisions}
fragment_ids <- which(dv$sentence_type == "fragment")
pred_ids     <- pmax(1L, fragment_ids - 1L)
pair_rows    <- sort(unique(c(pred_ids, fragment_ids)))

dv[pair_rows, ] |>
  mutate(
    role   = if_else(sentence_type == "fragment", "fragment", "predecessor"),
    reason = if_else(sentence_type == "fragment", reason, "")
  ) |>
  select(segid, role, reason, text) |>
  knitr::kable(
    col.names = c("Seg.", "Role", "Reason", "Text"),
    caption   = "Split decisions: fragments and the model's cited reason"
  )
```

### Coding decisions: near-cuts kept whole

Equally important are sentences the model *could* have split but correctly kept whole, applying the handbook's "when not to cut" rules. The manifesto contains several sentences with conjunctions or listed items that look splittable at first glance but express a single argument. Here are a handful of representative cases where the model's reason shows it considered and rejected a split:

```{r near-cuts}
near_cuts <- dv |>
  filter(
    sentence_type == "complete",
    grepl("\\band\\b|\\bor\\b", text),
    grepl(
      "single|elaborat|same|one (argument|claim|message|statement)|not.+(split|separate|unique|warrant)",
      reason, ignore.case = TRUE
    )
  ) |>
  slice_head(n = 5)

near_cuts |>
  select(segid, reason, text) |>
  knitr::kable(
    col.names = c("Seg.", "Reason", "Text"),
    caption   = "Near-cut decisions: sentences kept whole despite apparent complexity"
  )
```

## Inter-coder reliability of segmentation

How well does the LLM segmentation agree with the human gold standard? The `quasisentences` dataset includes the NZ National Party manifesto segmented by Manifesto Project coders. We can convert both the gold standard and the LLM output to segmented corpora and compare them with `qlm_compare()`, which computes Krippendorff's `_u_α` for unitizing — the standard reliability measure for segmented text (Krippendorff, 2019, section 12.6).

### Preparing the gold standard

The `quasisentences` dataset contains the Manifesto Project's human-coded quasi-sentences. We filter to the NZ manifesto, set a `docid` matching the LLM corpus, and convert to a segmented corpus with `as_qlm_coded()`.

```{r prepare-gold}
load("data/quasi-sentences/quasisentences.rda")

gold_nz <- quasisentences[quasisentences$manifesto == "NP 1972", ]
gold_nz$docid <- "NZ_NP_1972"

gold_corp <- as_qlm_coded(
  gold_nz,
  qlm_segment = TRUE,
  source_text = c(NZ_NP_1972 = manifesto_text),
  name        = "Manifesto Project",
  is_gold     = TRUE
)

gold_corp
```

### Comparing the segmentations

`qlm_compare()` detects that both inputs are segmented corpora and computes Krippendorff's `_u_α` for unitizing — the standard reliability measure for segmented text. Since `segs_manifesto` was produced by `qlm_segment()`, it already carries the character-level positions and metadata that `qlm_compare()` needs.

```{r compare-unitizations}
qlm_compare(segs_manifesto, gold_corp)
```

## Conclusion

`qlm_segment()` applies the Manifesto Project quasi-sentence rules to a raw manifesto, producing a quanteda corpus in which each document is a single political statement. The `reason` docvar makes it possible to audit both split and non-split decisions against the handbook rules. The segmented output can be compared to a human gold standard via `qlm_compare()`, which computes Krippendorff's `_u_α` for unitizing — measuring boundary agreement at the character level. The segmented corpus can also feed directly into `qlm_code()` for domain-level coding, reproducing the full Manifesto Project pipeline within a single R workflow.

## References

Burst, T., Krause, W., Lehmann, P., Lewandowski, J., Matthieß, T., Merz, N., Regel, S., Zehnter, L. (2021). *Manifesto Corpus. Version: South America 2021b*. Berlin: WZB Berlin Social Science Center.

Volkens, A., Burst, T., Krause, W., Lehmann, P., Matthieß, T., Merz, N., Regel, S., Weßels, B., Zehnter, L. (2021). *The Manifesto Data Collection. Manifesto Project (MRG/CMP/MARPOR). Version 2021b*. Berlin: WZB Berlin Social Science Center. https://doi.org/10.25522/manifesto.mpds.2021b
