---
title: "Example: Image analysis of campaign posters"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This example demonstrates multimodal analysis using `qlm_code()` to extract structured information from images. We analyze Indonesian mayoral campaign posters from Fox (2023), extracting details about candidates, visual elements, and symbolic content that would traditionally require manual human coding.

## Loading packages and data

```{r getting-started, message=FALSE}
library(quallmer)
library(dplyr)
library(knitr)
```

First, we identify the image files to analyze:

```{r identify-images}
# Get all image files from the data folder
image_files <- list.files("data/images/",
                          pattern = "\\.jpg$",
                          full.names = TRUE)

cat("Found", length(image_files), "campaign poster images:\n")
print(basename(image_files))
```

Let's preview one of the campaign posters:

```{r preview-image, echo=FALSE, out.width="60%", fig.align="center"}
knitr::include_graphics(image_files[1])
```

## Defining the image analysis codebook

We create a codebook that operationalizes the annotation task. The schema defines what information to extract from each poster:

```{r define-codebook}
# Define a comprehensive image analysis codebook
codebook_posters <- qlm_codebook(
  name = "Campaign Poster Analysis",
  instructions = paste(
    "You are a political scientist analyzing campaign posters from an",
    "Indonesian mayoral election. Examine each image carefully and",
    "extract the requested information about candidates, visual elements,",
    "and symbolic content."
  ),
  schema = ellmer::type_object(
    mayoral_candidate = ellmer::type_string(
      "Name of the mayoral candidate, or 'unknown' if not visible"
    ),
    deputy_candidate = ellmer::type_string(
      "Name of the deputy mayoral candidate, or 'unknown' if not visible"
    ),
    text_translation = ellmer::type_string(
      "English translation of the Indonesian text in the poster"
    ),
    clothing_description = ellmer::type_string(
      "Description of the clothing the candidates are wearing"
    ),
    indonesian_flag = ellmer::type_boolean(
      "Whether there are visual elements representing the red and white Indonesian flag"
    ),
    religious_buildings = ellmer::type_string(
      "Any religious buildings present and their religion, or 'none' if absent"
    ),
    party_logos = ellmer::type_boolean(
      "Whether there are any party logos in the poster"
    ),
    candidate_percentage = ellmer::type_integer(
      "Estimated percentage of poster taken up by faces and names of candidates (0-100)"
    ),
    facial_expression = ellmer::type_enum(
      c("smiling", "serious", "neutral", "mixed"),
      "Description of candidates' facial expressions"
    )
  ),
  role = "You are an expert in political communication and visual analysis.",
  input_type = "image"
)

# View the codebook structure
codebook_posters
```

The codebook includes:
- **Factual information**: Candidate names, text translations
- **Visual elements**: Clothing, flags, religious symbols, party logos
- **Compositional features**: Candidate prominence (percentage), facial expressions

The printed codebook also shows an image resize setting. Image files are resized before they are sent, and the resolution is part of the codebook because it is part of the measurement: names and small print that are legible at one size may not be at another, and a replication should read the image the original run read. The default, `image_file_resize = "high"`, fits each image within 2000x768 or 768x2000 pixels. `"low"` fits within 512x512, `"none"` sends the file as it is, and a magick geometry string such as `"1024x1024>"` sets a size of your own. Anything but `"none"` needs the magick package.

The results below were coded before this setting existed, at what was then the only resolution, `"low"`, and the saved codebook records that. Coding the posters again at `"high"` would let the model read the candidates' names and the small print in the party logos at full size, so the two runs may differ on those fields.

## Coding images using Gemini 3 Pro Preview

Multimodal models like Gemini 3 Pro Preview can analyze images and extract structured information. We use `qlm_code()` with image file paths:

```{r code-images, eval=FALSE}
# Apply image analysis using qlm_code()
coded_posters <- qlm_code(
  image_files,
  codebook = codebook_posters,
  model = "google_gemini/gemini-3-pro-preview",
  name = "campaign_posters_gemini3pro",
  notes = "Analysis of Indonesian mayoral campaign posters from Fox (2023)",
  include_cost = TRUE
)

# Add filenames to results
coded_posters$.filename <- basename(image_files)

# Save results
saveRDS(coded_posters, "data/coded_posters_gemini3pro.rds")
```

```{r load-results, echo=FALSE}
# Load pre-computed results
coded_posters <- readRDS("data/coded_posters_gemini3pro.rds")
coded_posters$.filename <- basename(image_files)
```

## Examining the results

Let's view the extracted information in a table:

```{r display-results}
# Display key results
coded_posters %>%
  select(.filename, mayoral_candidate, deputy_candidate, facial_expression,
         indonesian_flag, party_logos, candidate_percentage) %>%
  kable(
    col.names = c("File", "Mayoral Candidate", "Deputy", "Expression",
                  "Flag", "Logos", "% Candidates"),
    caption = "Campaign Poster Analysis Results"
  )
```

Total cost for analyzing `r length(image_files)` images:
(May not display correctly for a preview model)
```{r display-cost}
cat("Total cost: $", round(sum(coded_posters$cost, na.rm = TRUE), 4), sep = "")
```

### Text translations

The LLM can translate Indonesian text found in the posters:

```{r text-translations}
coded_posters %>%
  select(.filename, text_translation) %>%
  kable(
    col.names = c("File", "Text Translation"),
    caption = "Translated Text from Posters"
  )
```

### Visual elements

Summary of visual elements across all posters:

```{r visual-summary}
# Summarize visual elements
cat("Indonesian flag elements:",
    sum(coded_posters$indonesian_flag, na.rm = TRUE),
    "of", nrow(coded_posters), "posters\n")

cat("Party logos present:",
    sum(coded_posters$party_logos, na.rm = TRUE),
    "of", nrow(coded_posters), "posters\n")

cat("\nFacial expressions:\n")
print(table(coded_posters$facial_expression))

cat("\nCandidate prominence (% of poster):\n")
cat("Range:", min(coded_posters$candidate_percentage, na.rm = TRUE), "-",
    max(coded_posters$candidate_percentage, na.rm = TRUE), "%\n")
cat("Mean:", round(mean(coded_posters$candidate_percentage, na.rm = TRUE), 1), "%\n")
```

## Detailed view of one poster

Let's examine the complete analysis for one poster:

```{r detailed-view}
# Select the first poster for detailed view
poster_detail <- coded_posters[1, ]

cat("=== Detailed Analysis ===\n\n")
cat("File:", poster_detail$.filename, "\n\n")
cat("Mayoral candidate:", poster_detail$mayoral_candidate, "\n")
cat("Deputy candidate:", poster_detail$deputy_candidate, "\n")
cat("Text translation:", poster_detail$text_translation, "\n\n")
cat("Clothing:", poster_detail$clothing_description, "\n")
cat("Religious buildings:", poster_detail$religious_buildings, "\n\n")
cat("Indonesian flag present:", poster_detail$indonesian_flag, "\n")
cat("Party logos present:", poster_detail$party_logos, "\n")
cat("Candidate percentage:", poster_detail$candidate_percentage, "%\n")
cat("Facial expression:", poster_detail$facial_expression, "\n")
```

```{r show-detailed-image, echo=FALSE, out.width="60%", fig.align="center"}
knitr::include_graphics(image_files[1])
```

## Comparing to other models (optional)

You can code the same images with different models to compare results:

```{r compare-models, eval=FALSE}
# Try with GPT-4o for comparison
coded_gpt4o <- qlm_code(
  image_files,
  codebook = codebook_posters,
  model = "openai/gpt-4o",
  name = "campaign_posters_gpt4o"
)

# Compare agreement between models
qlm_compare(
  coded_posters,
  coded_gpt4o,
  by = "facial_expression",
  level = "nominal"
)
```

## Creating an audit trail

Document the complete analysis:

```{r trail, eval=FALSE}
qlm_trail(coded_posters, path = "poster_analysis")
```

This creates two files:

- `poster_analysis.rds`: Complete trail object containing the coding run, codebook, and metadata
- `poster_analysis.qmd`: Quarto document with full audit trail documentation

## Summary

This example demonstrates:

1. **Multimodal analysis**: Using vision-language models to analyze images
2. **Structured extraction**: Defining a schema to extract specific information
3. **Scalability**: Analyzing multiple images in batch
4. **Cost efficiency**: Modern multimodal models are increasingly affordable
5. **Reproducibility**: All analysis is documented and can be replicated

Multimodal LLMs open new possibilities for qualitative researchers working with visual data at scale, from political communication to social media analysis.

## References

Fox, C. A. (2023). Ethnic campaign appeals: To bond, bridge, or bypass? *Political Communication*, 40(1), 92–114. https://doi.org/10.1080/10584609.2022.2132331
