---
title: "Example: Audio transcription and analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This example demonstrates a two-step process for analyzing audio content: (1) transcribing audio to text using OpenAI's Whisper model, and (2) extracting structured information from the transcripts using `qlm_code()`. This workflow enables large-scale analysis of speeches, interviews, and other audio content.

## Loading packages and data

```{r getting-started, message=FALSE}
library(quallmer)
library(dplyr)
library(purrr)
library(knitr)
```

First, we identify the audio files to analyze:

```{r identify-audio}
# Get all audio files from the data folder
audio_files <- list.files("data/audio/",
                          pattern = "\\.(wav|mp3)$",
                          full.names = TRUE)

cat("Found", length(audio_files), "audio files:\n")
for (f in audio_files) {
  size_mb <- file.size(f) / 1024^2
  duration_sec <- NA  # Would require audio package to calculate
  cat(sprintf("  %s (%.2f MB)\n", basename(f), size_mb))
}
```

These audio files contain speech samples in multiple languages from the Open Speech Repository and other sources.

## Step 1: Transcribing audio with Whisper

OpenAI's Whisper model provides high-quality, multilingual transcription. We use the `openai` package to transcribe each audio file:

```{r transcribe-audio, eval=FALSE}
library(openai)

# Transcribe all audio files
transcriptions <- map_chr(audio_files, function(file_path) {
  cat("Transcribing:", basename(file_path), "\n")

  transcription <- create_transcription(
    file = file_path,
    model = "whisper-1"
  )

  transcription$text
})

# Name the transcriptions by filename
names(transcriptions) <- basename(audio_files)

# Save transcriptions
saveRDS(transcriptions, "data/transcriptions_whisper.rds")
```

```{r load-transcriptions, echo=FALSE}
# Load pre-computed transcriptions
transcriptions <- readRDS("data/transcriptions_whisper.rds")
```

## Viewing the transcriptions

Let's examine the transcribed content:

```{r view-transcriptions}
# Display each transcription (truncated for readability)
for (i in seq_along(transcriptions)) {
  cat("=== File:", names(transcriptions)[i], "===\n")

  # Show first 200 characters
  text_preview <- substr(transcriptions[i], 1, 200)
  if (nchar(transcriptions[i]) > 200) {
    text_preview <- paste0(text_preview, "...")
  }
  cat(text_preview, "\n\n")
}

# Word counts
word_counts <- map_int(transcriptions, ~length(strsplit(.x, "\\s+")[[1]]))
cat("Word count statistics:\n")
cat("  Range:", min(word_counts), "-", max(word_counts), "words\n")
cat("  Mean:", round(mean(word_counts)), "words\n")
```

## Step 2: Defining the transcript analysis codebook

Now we create a codebook to extract structured information from the transcripts. This codebook analyzes language, topics, tone, and content:

```{r define-codebook}
# Define a comprehensive transcript analysis codebook
codebook_transcripts <- qlm_codebook(
  name = "Speech Transcript Analysis",
  instructions = paste(
    "You are a research assistant analyzing transcribed speech content.",
    "Provide structured analysis of the speech based on its content.",
    "Be objective and accurate in your assessments."
  ),
  schema = ellmer::type_object(
    language = ellmer::type_string(
      "Primary language of the speech, in English (e.g., 'English', 'Mandarin', 'French', 'Indonesian')"
    ),
    language_confidence = ellmer::type_enum(
      c("high", "medium", "low"),
      "Confidence in language identification"
    ),
    speech_type = ellmer::type_enum(
      c("conversational", "formal_speech", "reading", "spontaneous", "other"),
      "Type or style of speech"
    ),
    main_topics = ellmer::type_string(
      "Main topics or themes discussed in the speech (comma-separated, max 5 topics)"
    ),
    key_phrases = ellmer::type_string(
      "Important phrases or keywords mentioned (comma-separated, max 5 phrases)"
    ),
    tone = ellmer::type_enum(
      c("formal", "informal", "neutral", "technical", "conversational"),
      "Overall tone of the speech"
    ),
    sentiment = ellmer::type_enum(
      c("positive", "negative", "neutral", "mixed"),
      "Overall sentiment expressed in the speech"
    ),
    summary = ellmer::type_string(
      "Brief 2-3 sentence summary of the speech content"
    )
  ),
  role = "You are an expert linguist and discourse analyst.",
  input_type = "text"
)

# View the codebook structure
codebook_transcripts
```

## Coding transcripts using Gemini 2.5 Flash

We use Gemini 2.5 Flash to analyze the transcripts. This model is fast and cost-effective for text analysis:

```{r code-transcripts, eval=FALSE}
# Apply transcript analysis using qlm_code()
coded_transcripts <- qlm_code(
  transcriptions,
  codebook = codebook_transcripts,
  model = "google_gemini/gemini-2.5-flash",
  name = "audio_transcripts_gemini",
  notes = "Analysis of multilingual speech transcripts",
  include_cost = TRUE
)

# Add filenames to results
coded_transcripts$.filename <- names(transcriptions)

# Save results
saveRDS(coded_transcripts, "data/coded_transcripts_gemini.rds")
```

```{r load-results, echo=FALSE}
# Load pre-computed results
coded_transcripts <- readRDS("data/coded_transcripts_gemini.rds")
coded_transcripts$.filename <- names(transcriptions)
```

## Examining the results

Let's view the extracted information:

```{r display-results}
# Display key results
coded_transcripts %>%
  select(.filename, language, language_confidence, speech_type,
         tone, sentiment) %>%
  kable(
    col.names = c("File", "Language", "Confidence", "Type", "Tone", "Sentiment"),
    caption = "Transcript Analysis Results"
  )
```

Total cost for analyzing `r length(transcriptions)` transcripts:
```{r display-cost}
cat("Transcription cost (Whisper): ~$",
    round(sum(word_counts) * 0.006 / 100, 4), " (estimated)\n", sep = "")
cat("Analysis cost (Gemini): $",
    round(sum(coded_transcripts$cost, na.rm = TRUE), 4), "\n", sep = "")
```

### Language distribution

```{r language-distribution}
cat("Languages detected:\n")
language_table <- table(coded_transcripts$language)
print(language_table)

cat("\nLanguage confidence levels:\n")
print(table(coded_transcripts$language_confidence))
```

### Topics and themes

```{r topics}
# Display topics for each transcript
coded_transcripts %>%
  select(.filename, main_topics) %>%
  kable(
    col.names = c("File", "Main Topics"),
    caption = "Topics Identified in Each Transcript"
  )
```

### Detailed view of one transcript

Let's examine the complete analysis for one transcript:

```{r detailed-view}
# Select the first transcript for detailed view
transcript_detail <- coded_transcripts[1, ]

cat("=== Detailed Analysis ===\n\n")
cat("File:", transcript_detail$.filename, "\n\n")
cat("Language:", transcript_detail$language,
    "(", transcript_detail$language_confidence, "confidence )\n")
cat("Speech type:", transcript_detail$speech_type, "\n")
cat("Tone:", transcript_detail$tone, "\n")
cat("Sentiment:", transcript_detail$sentiment, "\n\n")
cat("Main topics:", transcript_detail$main_topics, "\n\n")
cat("Key phrases:", transcript_detail$key_phrases, "\n\n")
cat("Summary:", transcript_detail$summary, "\n\n")

cat("=== Original Transcription (first 300 chars) ===\n")
cat(substr(transcriptions[1], 1, 300), "...\n")
```

## Summary statistics

```{r summary-stats}
# Speech type distribution
cat("Speech types:\n")
print(table(coded_transcripts$speech_type))

# Tone distribution
cat("\nTone distribution:\n")
print(table(coded_transcripts$tone))

# Sentiment distribution
cat("\nSentiment distribution:\n")
print(table(coded_transcripts$sentiment))
```

## Creating an audit trail

Document the complete analysis:

```{r trail, eval=FALSE}
qlm_trail(coded_transcripts, path = "audio_analysis")
```

This creates two files:

- `audio_analysis.rds`: Complete trail object containing the coding run, codebook, and metadata
- `audio_analysis.qmd`: Quarto document with full audit trail documentation

## Summary

This example demonstrates:

1. **Two-step workflow**: Transcription (Whisper) → Analysis (LLM)
2. **Multilingual capability**: Whisper handles multiple languages automatically
3. **Structured extraction**: Codebooks define what to extract from transcripts
4. **Scalability**: Process multiple audio files in batch
5. **Cost efficiency**: Whisper is significantly cheaper than human transcription
6. **Reproducibility**: All steps are documented and can be replicated

This workflow enables researchers to analyze audio content at scale, from political speeches to interviews to podcast episodes. The combination of Whisper for transcription and LLMs for analysis provides both accuracy and interpretability.
