---
title: "Example: Text segmentation"
subtitle: "Aspect-based sentiment and thematic analysis with `qlm_segment()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example: Text segmentation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5
)
```

`qlm_segment()` splits a text into thematic or conceptual units and returns a [quanteda corpus](https://quanteda.io/) with one document per segment. The schema fields from the codebook become docvars on the output corpus, making it straightforward to combine LLM-powered segmentation with downstream quantitative analysis.

This article illustrates two applications: aspect-based sentiment analysis of film reviews, and thematic segmentation of recent US presidential inaugural addresses.

## Packages

```{r packages, message=FALSE, warning=FALSE}
library(quallmer)
library(quanteda)
library(dplyr)
library(ggplot2)
```

## Aspect-based sentiment in movie reviews

Aspect-based sentiment analysis (ABSA) asks not just whether a reviewer is positive or negative overall, but which specific features they discuss and what they say about each. A review that praises the acting while criticising the plot will read as moderately positive in aggregate, but the aspect-level picture is much richer.

### The data

We draw four reviews from `data_corpus_LMRDsample`, the 200-review subset of the Large Movie Review Dataset bundled with quallmer (Maas et al., 2011): two negative and two positive.

```{r sample-reviews}
set.seed(55)
idx_neg <- which(docvars(data_corpus_LMRDsample)$polarity == "neg")
idx_pos <- which(docvars(data_corpus_LMRDsample)$polarity == "pos")
corp_reviews <- data_corpus_LMRDsample[c(sample(idx_neg, 2), sample(idx_pos, 2))]
corp_reviews
```

### The codebook

```{r codebook-absa}
cb_absa <- qlm_codebook(
  name = "Aspect-based sentiment",
  instructions = paste(
    "Segment the review into contiguous spans that each address a single aspect",
    "of the film. A segment may be a clause, a sentence, or several sentences.",
    "Label each segment with the aspect it discusses and the expressed sentiment.",
    "",
    "Aspects: acting, cinematography, direction, music, pacing, plot, writing, other.",
    "Return the verbatim text of each segment."
  ),
  schema = ellmer::type_object(
    aspect    = ellmer::type_enum(
      c("acting", "cinematography", "direction", "music", "pacing",
        "plot", "writing", "other"),
      description = "Film aspect discussed in this segment"
    ),
    sentiment = ellmer::type_enum(
      c("negative", "neutral", "positive"),
      description = "Sentiment expressed toward this aspect"
    )
  ),
  role = "You are an expert film critic and NLP researcher."
)

cb_absa
```

### Segmenting the reviews

```{r segment-reviews, eval=FALSE}
segs_reviews <- qlm_segment(
  corp_reviews,
  codebook = cb_absa,
  model    = "openai/gpt-4o-mini",
  name     = "GPT-4o-mini"
)
saveRDS(segs_reviews, "data/segs_reviews.rds")
```

```{r load-segs-reviews, echo=FALSE}
segs_reviews <- readRDS("data/segs_reviews.rds")
```

### Segments

Each review is now split into aspect-level segments. The output corpus carries `docid`, `segid`, `aspect`, and `sentiment`, plus the `polarity` and `rating` docvars inherited from the source corpus:

```{r show-reviews}
docvars(segs_reviews) |>
  mutate(text = as.character(segs_reviews)) |>
  select(docid, segid, aspect, sentiment, polarity, rating, text) |>
  knitr::kable()
```

### Sentiment by aspect

Pooling across all four reviews, we can see which aspects attract positive versus negative commentary:

```{r plot-absa}
docvars(segs_reviews) |>
  count(aspect, sentiment) |>
  mutate(
    sentiment = factor(sentiment, levels = c("negative", "neutral", "positive")),
    aspect    = reorder(aspect, n, sum)
  ) |>
  ggplot(aes(x = aspect, y = n, fill = sentiment)) +
  geom_col() +
  scale_fill_manual(
    values = c(negative = "#d9534f", neutral = "#f0ad4e", positive = "#5cb85c")
  ) +
  labs(
    x     = NULL,
    y     = "Number of segments",
    title = "Aspect-based sentiment across four film reviews",
    fill  = "Sentiment"
  ) +
  coord_flip() +
  theme_minimal()
```

## Thematic segmentation of inaugural addresses

Presidential inaugural addresses typically move through several distinct themes — calls for unity, foreign policy commitments, economic priorities, statements of national values — often without explicit markers separating them. `qlm_segment()` can recover this thematic structure automatically.

### The data

We use the two most recent speeches in `quanteda::data_corpus_inaugural`:

```{r load-inaugural}
corp_inaugural <- tail(data_corpus_inaugural, 2)
corp_inaugural
```

### The codebook

```{r codebook-inaugural}
cb_inaugural <- qlm_codebook(
  name = "Thematic segmentation of inaugural addresses",
  instructions = paste(
    "Segment this inaugural address into contiguous thematic passages.",
    "Begin a new segment when the speaker shifts to a clearly different theme.",
    "Each segment must contain at least one complete sentence.",
    "",
    "Themes:",
    "  democracy_governance  -- democratic institutions, rule of law, elections",
    "  economy_prosperity    -- jobs, industry, trade, economic growth",
    "  foreign_policy        -- international relations, alliances, diplomacy",
    "  military_security     -- armed forces, national security, veterans",
    "  national_unity        -- calls for solidarity, healing, shared identity",
    "  religion_values       -- religious references, moral and cultural values",
    "  social_issues         -- healthcare, education, civil rights, environment",
    "  other                 -- passages that do not fit the above themes"
  ),
  schema = ellmer::type_object(
    theme = ellmer::type_enum(
      c("democracy_governance", "economy_prosperity", "foreign_policy",
        "military_security", "national_unity", "religion_values",
        "social_issues", "other"),
      description = "Thematic label for this passage"
    )
  ),
  role = "You are an expert in American political rhetoric and speech analysis."
)

cb_inaugural
```

### Segmenting the speeches

```{r segment-inaugural, eval=FALSE}
segs_inaugural <- qlm_segment(
  corp_inaugural,
  codebook = cb_inaugural,
  model    = "openai/gpt-4.1-mini"
)
saveRDS(segs_inaugural, "data/segs_inaugural.rds")
```

```{r load-segs-inaugural, echo=FALSE}
segs_inaugural <- readRDS("data/segs_inaugural.rds")
```

### Summary of themes

```{r theme-table}
docvars(segs_inaugural) |>
  count(docid, theme) |>
  tidyr::pivot_wider(names_from = docid, values_from = n, values_fill = 0L) |>
  knitr::kable(caption = "Number of segments by theme and speech")
```

### Theme distribution by speech

```{r plot-inaugural, fig.height=5, fig.width=9}
docvars(segs_inaugural) |>
  count(docid, theme) |>
  mutate(theme = tools::toTitleCase(gsub("_", " ", theme))) |>
  ggplot(aes(x = reorder(theme, n, sum), y = n, fill = docid)) +
  geom_col(position = "dodge") +
  labs(
    x     = NULL,
    y     = "Number of segments",
    fill  = "Speech",
    title = "Thematic composition of two inaugural addresses"
  ) +
  coord_flip() +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Reliability of aspect-based segmentation

How reproducible is the ABSA segmentation? We re-segment the same four reviews with a second model and compare the two runs using `qlm_compare()`, which computes Krippendorff's `_u_α` for unitizing (Krippendorff, 2019, section 12.6).

### A second segmentation

```{r segment-reviews-2, eval=FALSE}
segs_reviews_2 <- qlm_segment(
  corp_reviews,
  codebook = cb_absa,
  model    = "anthropic/claude-sonnet-4-5",
  name     = "Claude Sonnet 4.5"
)
saveRDS(segs_reviews_2, "data/segs_reviews_2.rds")
```

```{r load-segs-reviews-2, echo=FALSE}
segs_reviews_2 <- readRDS("data/segs_reviews_2.rds")
```

### Boundary agreement

`|_u_α_binary` tests whether the two models place segment boundaries in the same locations, ignoring the aspect and sentiment labels:

```{r compare-binary}
qlm_compare(segs_reviews, segs_reviews_2)
```

### Boundary and sentiment agreement

When we pass `by = "sentiment"`, `_u_α_nominal` tests both boundary placement and whether the models agree on the sentiment label assigned to each span:

```{r compare-nominal}
qlm_compare(segs_reviews, segs_reviews_2, by = "sentiment")
```

The per-review breakdown shows which texts are more or less consistently segmented, while the overall row is the value to report.

## Conclusion

`qlm_segment()` turns unstructured text into a quanteda corpus where each document is a semantically coherent unit. The output integrates directly with quanteda's ecosystem — frequency tables, readability measures, or further LLM coding of individual segments with `qlm_code()`. The ABSA example shows how mixed sentiment can be obscured in aggregate scores; the inaugural address example shows how thematic structure can be recovered automatically from long-form political text. Segmented corpora can be compared across models or runs with `qlm_compare()`, which computes Krippendorff's `_u_α` for unitizing — measuring both boundary and coding agreement at the character level.

## References

Maas, A. L., Daly, R. E., Pham, P. T., Huang, D., Ng, A. Y., & Potts, C. (2011). Learning word vectors for sentiment analysis. *Proceedings of the 49th Annual Meeting of the Association for Computational Linguistics*, 142–150.
