---
title: "Example: Fact checking of claims"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The `annotate()` function with a predefined `task_fact()` can be used to fact-check claims made in texts. In this example, we will demonstrate how to apply this task to a sample corpus of innaugural speeches from US presidents.

### Loading packages and data

```{r getting-started}
# We will use the quanteda package 
# for loading a sample corpus of innaugural speeches
# If you have not yet installed the quanteda package, you can do so by:
# install.packages("quanteda")
library(quanteda)
library(quallmer)

# For educational purposes, 
# we will use a subset of the inaugural speeches corpus
# The three most recent speeches in the corpus
data_corpus_inaugural <- quanteda::data_corpus_inaugural[57:60]

```
### Using `annotate()` for fact checking of claims in texts
```{r fact}

# Apply predefined fact checking task with task_fact() in the annotate() function
result <- annotate(data_corpus_inaugural, task = task_fact(), 
                   chat_fn = chat_openai, model = "gpt-4o",
                   api_args = list(temperature = 0, seed = 42))
```

```{r display_results, echo = FALSE}
# results table
library(kableExtra)
library(dplyr)
result <- result %>%
  dplyr::select(id, truth_score, misleading, explanation)
result %>%
  kable("html", escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  column_spec(3)
```
### Using `annotate()` for fact checking with a specific number of claims to check
```{r fact2}

# Apply predefined fact checking task with task_fact() in the annotate() function
result_claims <- annotate(data_corpus_inaugural, task = task_fact(max_topics = 3), 
                   chat_fn = chat_openai, model = "gpt-4o",
                   api_args = list(temperature = 0, seed = 42))
```
```{r display_results2, echo = FALSE}
# results table
library(kableExtra)
library(dplyr)
result_claims <- result_claims %>%
  dplyr::select(id, truth_score, misleading, explanation)
result_claims %>%
  kable("html", escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  column_spec(3)
```
