---
title: "Example: Thematic coding"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The `annotate()` function with a predefined sentiment task allows you to rate the sentiment of texts using an LLM. The predefined `themes()` object structures the response by indicating whether specific themes are present in the text, along with proportion scores (0-1) and explanations for each theme.

### Loading packages and data

```{r getting-started}
# We will use the quanteda package for loading a sample corpus of innaugural speeches
# If you have not yet installed the quanteda package, you can do so by:
# install.packages("quanteda")
library(quanteda)
library(quallmer)

# For educational purposes, we will use a subset of the inaugural speeches corpus
# The three most recent speeches in the corpus
data_corpus_inaugural <- quanteda::data_corpus_inaugural[57:60]

```

### Using `annotate()` for thematic coding of texts

```{r themes}
# Define themes of interest
topics <- c("Politics", "Economy", "Artificial Intelligence", "Environment", "Healthcare")
# Apply predefined themes task with themes() in the annotate() function
result <- annotate(data_corpus_inaugural, task = themes(topics), 
                   chat_fn = chat_openai, model = "gpt-4o",
                   api_args = list(temperature = 0, seed = 42))
```
```{r display_results, echo = FALSE}
# thematic coding results table
library(kableExtra)
result %>%
  kable("html", escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  column_spec(3)
```
